Self Referencing Sequences

### 2014-10-28 draft.tex
Edits to Iterative Process and Substitution Rules

One possible iterative process is as follows. Let the sequence $\{s^0\}$ be 2, and let the sequence $\{s^{(i)}\}$ for $i \geq 1$ be the unique sequence that starts with 2 and is such that its block lengths, read from left to right, reproduce the sequence $\{s^{(i-1)}\}$. This gives:
\begin{align*}
\{s^{(0)}\} &= 2\\
\{s^{(1)}\} &= 2,2\\
\{s^{(2)}\} &= 2,2,1,1\\
\{s^{(3)}\} &= 2,2,1,1,2,1\\
\{s^{(4)}\} &= 2,2,1,1,2,1,2,2,1\\
&\cdots
\end{align*}
Let $s^\infty$ be the limit of this process. By limit, we mean the limit with respect to finitely many terms $s^{(i)}_j \rightarrow s^\infty_j$ as$i \rightarrow \infty$. We claim that $s^\infty$ is the unique self-referential sequence $s(\{1,2\},2)$ generated by $\{1,2\}$ with starting number 2.
\begin{proof}
We prove this with induction. We show that if $s^{(i)}$ is a prefix of $s(\{1,2\},2)$, then $s^{(i+1)}$ is a longer prefix of $s(\{1,2\},2)$. The base case of the induction is satisfied, as $s^{(0)}$ is a prefix of $s(\{1,2\},2)$.

For the inductive step, observe that a prefix of the sequence describes the block lengths of a \emph{longer} prefix of the sequence. We know that it describes a longer prefix because the prefix it describes has length equal to the sum of the original prefix. In other words, the length of $s^{(i+1)}$ must be equal to the sum of of the numbers in $s^{(i)}$, which is strictly greater than the length of $s^{(i)}$ (since there are numbers that are greater than 1). This holds true because $s^{(i)}$ merely contains a collection of lengths, which are summed up to give the next prefix. 
\end{proof}

It turns out that this same argument holds in general.

\begin{theorem}
Let $\{s^0\} = a$ for some $a \in A$ where $a \neq 1$, and let $\{s^{(i)}\}$ for $i \geq 1$ be the unique sequence that starts with $a$ and is such that its block lengths, read from left to right, reproduce the sequence $\{s^{(i-1)}\}$. Then the limit of this process, $s^\infty$, is the sequence $s(A,a)$.
\end{theorem}

Note that we must specify that $a \neq 1$, because that is the only case in which the length of $s^{(i+1)}$ is not strictly greater than the length of $s^{(i)}$. The proof of this theorem is omitted, as it is essentially the same proof as for the $A = \{1,2\}, a = 2$ case.

It turns out that we can be less restrictive about what the starting sequence is. Going back to the case of $\{1,2\}$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2. For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating large $s^{(i)}$ give:
\begin{align*}
\{s^{(0)}\} &= \mathbf{2},1,1,1,2\\
\{s^{(1)}\} &= \mathbf{2,2,1},2,1,2,2\\
\{s^{(2)}\} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
\{s^{(3)}\} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\cdots
\end{align*}
Here, we highlight in bold the numbers that form a prefix of $s(\{1,2\},2)$. We note that $s^\infty = s(A,a)$ simply because $s^{(0)}$ contains a 2 at the beginning, whch then causes a 2,2 to appear at the start of $s^{(1)}$, which causes $2,2,1,1$ to appear at the start of $s^{(2)}$, and so on. The rest of each $s^{(i)}$ that is not a prefix of $s({1,2},2)$ we call the \emph{tail}.

We now consider two different forms of convergence. We know that $s^{(i)}$ in some sense approaches $s(A,a)$, as arbitrarily many elements at the beginning of the sequence will match up with the beginning of $s(A,a)$ as $i$ increases. However, if the tail of the sequence never entirely disappears, then there does not exist an $i$ where $s^{(i)}$ is exactly a prefix of $s(A,a)$. Thus, we can formalize two types of convergence: \emph{weak convergence} and \emph{strong convergence}. Weak convergence is satisfied if $s^\infty = s(A,a)$, and strong convergence is satisfied if $s^{(i)}$ is a prefix of $s(A,a)$ for all $i \geq k$ for some $k$.

@@

So we can ask, when can we do this? We would like to be able to simulate our original algorithm with this current one. We had established that our algorithm had a state corresponding to 
\begin{itemize}
\item current block length number
\item how many elements already written down in the block
\item which number we need to write.
\end{itemize}

In an algorithm with rules, we cannot remember anything about the previous rules. Therefore, each rule needs to encompas the block length numbers, the number of elements already written, and which number we need to write. The first condition is trivial to keep, since the rules themselves will say the block length numbers, in order to keep track of the numbers already written, each rule should be self-contained. That means that the rules should read off the block numbers. We would also need the length of the rules to be divisible by the size of the generating set. This is so that at the start of each rule, we always know which numbers to write.

Therefore, we can formulate a set of sufficient conditions to make rules to generate the self-referential sequence with generating set $G$. We want to have a set of rules $R$ where each rule $r \in R$ has length $|r| = l$, such that all $|G|^{l}$ strings are in the rule set and for each $r \in R$
\begin{itemize}
\item each rule reads off the block lengths in the same order of $G$
\item $\sum_{i \in r} i$ is divisible by $l$
\item $l$ is divisible by $|G|$
\end{itemize}
Given these conditions on every rule, we can guarantee that the self-referential sequence is generated. Note that in this case, the above example does not satisfy our rules until we say that $1, 3 \to 3, 1, 1, 1$.

Another exaple, we can have $G = \{ 2, 4 \}$, and we can have that the rules be
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}

>>>>

One possible iterative process is as follows. Let the sequence $s^0$ be 2, and let the sequence $s^{(i)}$ for $i \geq 1$ be the unique sequence that starts with 2 and is such that its block lengths, read from left to right, reproduce the sequence $s^{(i-1)}$. This gives:
\begin{align*}
s^{(0)}&= 2\\
s^{(1)} &= 22\\
s^{(2)} &= 2211\\
s^{(3)}&= 221121\\
s^{(4)} &= 221121221\\
&\vdots
\end{align*}
Let $s^\infty$ be the limit of this process. 
By limit, we mean the limit with respect to finitely many terms $s^{(i)}_j \rightarrow s^\infty_j$ as $i \rightarrow \infty$. We claim that $s^\infty$ is the unique self-referential sequence $s(\{1,2\},2)$ generated by $\{1,2\}$ with starting number 2. In fact we can prove a more general result about this iterative process.

\begin{theorem}
Let $s^0 = a$ for some $a \in A$ where $a \neq 1$, and let $s^{(i)}$ for $i \geq 1$ be the unique sequence that starts with $a$ and is such that its block lengths, read from left to right, reproduce the sequence $s^{(i-1)}$. Then the limit of this process, $s^\infty$, is the sequence $s(A,a)$.
\end{theorem}

\begin{proof}
We prove this with induction. We show that if $s^{(i)}$ is a prefix of $s(A,a)$, then $s^{(i+1)}$ is a longer prefix of $s(A,a)$. The base case of the induction is satisfied, as $s^{(0)}$ is a prefix of $s(A,a)$.

For the inductive step, we observe that a prefix of the sequence describes the block lengths of a \emph{longer} prefix of the sequence. We know that the sequence it describes is longer because the sequence it describes has length equal to the sum of the original prefix. In other words, the length of $s^{(i+1)}$ equals the sum of of the numbers in $s^{(i)}$, which is strictly greater than the length of $s^{(i)}$ (since there are numbers that are greater than 1).

The resultant sequence is a prefix of the self referencing sequence because if we apply our process to any prefix of the self referencing sequence it will produce a prefix of the self referencing sequence. 
\end{proof}

Note that we must specify that $a \neq 1$, because that is the only case in which the length of $s^{(i+1)}$ is not strictly greater than the length of $s^{(i)}$.

It turns out that we can be less restrictive about what our starting sequence is. Going back to the case of $\{1,2\}$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2. For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= \mathbf{2},1,1,1,2\\
s^{(1)} &= \mathbf{2,2,1},2,1,2,2\\
s^{(2)} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
s^{(3)} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
Here, we bold the numbers that form a prefix of $s(\{1,2\},2)$. We note that $s^\infty = s(A,a)$ because because $s^{(0)}$ contains a 2 at the beginning, which by the above theorem will produce $s(A,a)$ in the limit. We call the rest of each $s^{(i)}$ that is not a prefix of $s({1,2},2)$ the \emph{tail}.

We now consider two different forms of convergence. We know that $s^{(i)}$ in some sense approaches $s(A,a)$, as arbitrarily many elements at the beginning of the sequence will match up with the beginning of $s(A,a)$ as $i$ increases. However, if the tail of the sequence never entirely disappears, then there does not exist an $i$ where $s^{(i)}$ is exactly a prefix of $s(A,a)$. Thus, we formalize two types of convergence
\begin{definition} We say that a sequence $s^{i}$ is weakly convergent to $s(A,a)$ if $s^\infty = s(A,a)$. We say that the sequence $s^{i}$ is strongly convergent if $s^{(i)}$ is a prefix of $s(A,a)$ for all $i \geq k$ for some $k$.
\end{definition}
It is unclear what conditions must be placed on the starting sequence in order to ensure strong convergence, although based on numerical simulation we make the following conjecture.

\begin{conjecture} A sequence $s^i$ is strongly convergent to $s(\{1,2\}, 2)$ if the starting sequence begins with a $2$ and contains no more than 3 elements. A sequence $s^i$ is strongly convergent to $s(\{1,2\}, 1)$ if the starting sequence begins with a $1$ and contains between 2 and 3 elements. A sequence $s^i$ is weakly convergent but not strongly convergent to $s(\{1,2\}, a)$ if the starting sequence begins with an $a \in \{1,2,\}$ and contains 4 or more elements.
\end{conjecture}
We make this conjecture on the basis on computer simulations which executes the iterative process to at least 20 iterations. More iterations were not completed because after this point the simulations began to take impractically long amounts of time to complete.

@@

We would like to know if we can do this for any generating set $G$ and starting number $g$. In particular we would like to be able to simulate our original algorithm with a substitution rule. Previously we established that our algorithm had a state corresponding to 
\begin{itemize}
\item current block size
\item number of elements already written down in the current block
\item what number the current block is composed of 
\end{itemize}

When we apply our substitution rules we should not need to know what happened to blocks other than the one we are currently substituting for. Therefore, each rule must somehow encode the length of  blocks, the number of elements already written in the current block, and what number the current block is composed of. The first condition is simple to implement, as the rules will specify the length of blocks. In order to keep track of the numbers already written, each rule should be self-contained. That is each rule should write complete blocks. In order to comply with the third condition it is enough for the size of the generating set to divide the length of the rules. This is so that at the start of each rule, we always know which numbers to write.

Therefore, we can formulate a set of sufficient conditions to make rules to generate the self-referential sequence with generating set $G$. We want to have a set of rules $R$ where each rule $r \in R$ has length $|r| = l$, such that all $|G|^{l}$ strings are in the rule set and for each $r \in R$
\begin{itemize}
\item each rule writes blocks in the same order of $G$
\item $\sum_{i \in r} i$ is divisible by $l$
\item $l$ is divisible by $|G|$
\end{itemize}
Given these conditions on every rule, we can guarantee that the self-referential sequence is generated. Note that in this case, the above example does not satisfy our rules until we say that $1, 3 \to 3, 1, 1, 1$.

Another example, we can have $G = \{ 2, 4 \}$, and we can have that the rules be
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}
